## Working with Orthoimagery

### Load necessary libraries

Let's load some basic libraries to work with the data

```{r}
#| echo: true
#| output: false
renv::restore()
library(here)
library(dplyr)
library(ggplot2)
library(sf)
library(terra)
library(vapour)
```

### Initialize metadata catalog

The data catalog is basically a geojson file which is stored in a cloud bucket. It consists of many polygons which cover cover Lower Saxony and represent image tiles (2x2 km). For each tile it contains some basic metadata and links to further metadata of the images and links to the image data itself.

Let's load the data in R.

```{r}
#| echo: true
dat <- sf::st_read("https://single-datasets.opengeodata.lgln.niedersachsen.de/pro-download-indices/dop/lgln-opengeodata-dop20.geojson")

```

and print the first object (tile) to get an ideas of the data.

```{r}
#| echo: true
print(dat[1,])
```

we can optionally preprocess the data, here we change the date format

```{r}
#| echo: true
dat$date <- format(as.Date(as.character(dat$Aktualitaet), format = "%Y-%m-%d"), format= "%Y%m%d")
dat$year <- format(as.Date(as.character(dat$Aktualitaet), format = "%Y-%m-%d"), format= "%Y")
dat$month_day <- format(as.Date(as.character(dat$Aktualitaet), format = "%Y-%m-%d"), format= "%m%d")
dat$doy <- format(as.Date(as.character(dat$Aktualitaet), format = "%Y-%m-%d"), format= "%j")
```

### Plot metadata catalog

As we have seen above, the data catalog contains information about the acquisition date of images. We can plot this data to get more information about the available images.

#### Latest acquisition date per tile

```{r}
#| label: fig-date-latest
#| fig-cap: Latest acquisition date per tile
#| warning: false
#| echo: true
dat |> 
  arrange(date) |> 
  
  ggplot() + 
  geom_sf( aes(fill = year), color = NA) + 
  theme_void()  +
  scale_fill_viridis_d(option = "plasma")

```

#### Acquisition date within the years

```{r}
#| label: fig-date-doy
#| fig-cap: Acquisition date within year
#| warning: false
#| echo: true
dat |>  
  arrange(date) |> 
  
  ggplot() + 
  geom_sf( aes(fill = as.integer(doy)), color = NA) + 
  theme_void()  +
  scale_fill_viridis_c(option = "plasma")

# interactive mapping

# dat |> 
#   arrange(date) |>  
#   mapview::mapview(zcol="doy")
```

### Plot image data

Finally we can also query the image data itself. As we have seen above the data catalog contains links to the images. For each tile there is two different image versions per date: *rgb-images* are 3-band images compressed as jpeg and primarily meant to display; *rgbi-images* are lossless compressed and contain the NIR-channel, they can be used for analytical work. The images are stored in a cloud bucket as Cloud Optimized Geotiffs (COG), this means that the image data contains image overviews and is structured internally in such a way that it is possible to stream only the required image data. This makes it quite efficient and comfortable to work with the image data.

#### Latest image

Here we sort the metadata catalog in descending order by acquisition date and get the first rgb-image link. We then load and plot the data.

```{r}
#| label: fig-latest-image
#| fig-cap: Latest acquired image
#| warning: false
#| echo: true
fileadress =
  dat |> 
  arrange(desc(date)) |> 
  slice(1) |> 
  pull(rgb)

cog.url <- paste0("/vsicurl/", fileadress)
ras <- terra::rast(cog.url)

terra::plotRGB(ras)
```

#### First image

Here we sort the metadata catalog in ascending order by acquisition date and get the first rgbi-image link. We then load and plot the data as false color composite.

```{r}
#| label: fig-first-image
#| fig-cap: First acquired image in false color
#| warning: false
#| echo: true
fileadress =
  dat |> 
  arrange(date) |> 
  slice(1) |> 
  pull(rgbi)

cog.url <- paste0("/vsicurl/", fileadress)
ras <- terra::rast(cog.url)


terra::plotRGB(ras, r=4,g=3,b=2)

```

#### Multi-temporal

As mentioned before we can access just the data we need from COGs, this means we can query a certain extent or query the image with certain dimensions. In the examples above we plotted the overview, but in order to access the raw data we need some function in R. We create this function here with the vapor package which is able to warp the underlying raw data.

```{r}
#| echo: true
#| output: false
get_img <- function(img, dim = c(500,500)){
  
  fileadress <- img |> 
    pull(rgb)  
  
  cog.url <- file.path("/vsicurl", fileadress)
  info <- vapour::vapour_raster_info(cog.url)
  roi <- info$extent
  prj <- info$projection
  dim <-  dim

  vals <- vapour::vapour_warp_raster(cog.url, extent = roi, dimension = dim, projection = prj
                                   , bands = c(1,2,3)
                                   , band_output_type = "Int32"
                                   , resample = "Bilinear")
  
  ras <- terra::rast(terra::ext(roi), ncols = dim[1], nrows = dim[2], vals = array(unlist(vals), c(dim, 3)), nlyrs = 3, crs = prj)

  return(ras)
}

```

Lets get all images for a certain tile and print the acquisition dates.

```{r}
#| echo: true
multitemporal <- dat |> 
  filter(tile_id == "324965790")

multitemporal$Aktualitaet
```

#### First image in tile

Now lets query and print the first image for this tile.

```{r}
#| label: fig-first-image-multitemporal
#| fig-cap: First image in tile
#| warning: false
#| echo: true
img1 <- multitemporal |> 
  slice_min(order_by = date) |> 
  get_img()

terra::plotRGB(img1)
```

#### Last image in tile

And the last image for this tile.

```{r}
#| label: fig-last-image-multitemporal
#| fig-cap: Last image in tile
#| warning: false
#| echo: true
img2 <- multitemporal |> 
  slice_max(order_by = date) |> 
  get_img()

terra::plotRGB(img2)
```
